### 2024-08-27 Chapters/Chapter4.tex
Updated after new CARRA data sourced and run

\begin{table}[h]
    \centering
    \caption{Hyperparamter search with best performing combination colored red}
    \label{table:gridSearchHyperparamters}
    \begin{tabular}{c|c}
        Layers &  4, \textcolor{red}{5}, 6, 7, 8, 9\\\hline
        Units &  128, 256, 512\\\hline
        Epochs & 100, 200, 500\\\hline
        Batch size & 32, 64, 128, 256\\\hline
        Optimizers & Adam, RMSprop, Adamax\\\hline
        Penalties & 1, 0.1, 0.01, 0\\\hline
    \end{tabular}
\end{table}

\section{Feature selection and importance}
Using the three datasources, a model is trained. The data from IMO only represents the ground truth and is not used as part of the training data. The gust ($f_g$) and wind speed ($f$) are used to calculate the target gust factor. All weather parameters used in predictions come from the CARRA reanalysis data, with variables wind speed, wind direction, temperature, pressure. All of these variables are queried in three height levels (15, 250, 500 meters above ground) at the location of a given weather station. The second data source for the training data, is the elevation data. Using the GeoTif file that IMO provided, several different landscape elevation distributions were looked at. A sector upwind, a sector upwind and downwind as well as a circle surrounding a weather station.

To begin with all the available parameters were used to train the model along with the derived variables. This is done to be able to then use tools such as Shapley to see which features are impacting the predictions. Using Shapley values to see which features, will then enable the exclusion of features that don't affect the model prediction and simplify the training data. This means starting with much higher dimensionality of our training data and reduce it as much as possible without impacting our results. This crystallizes in the landscape points. Starting with $n$ points that describe the elevation of the landscape around our weather station. This $n$ might be in the hundreds. Looking at concentric circles outward from a given weather station, there might be 10 circles (largest with radius 20 km), each with 72 points (5° spacing) and end up with 720 points for the landscape. This might slow down training. In addition, certain landscape features might be more important than others. There might be a large mountain up- or downwind of a given weather station, that might be the most important feature. To address this principal component analysis (PCA) is used to try to reduce the dimensionality of our landscape elevation data. PCA is a statistical procedure that allows the summarization of multivariate data to lower dimension and thus can ease visualization or increase speed of training a complex model with only minimal adverse effect to performance\cite{pca_information}. Another thing that might serve the same purpose is to select some number of $m$ points that have the most increase or decrease in elevation from the weather station along with the coordinates of those points in relation to the weather station and the direction of the wind.

Using all the variables from CARRA, along with the two derived factors ($Ri$, $N$) and two sectors (upwind and downwind) downscaled to 10 features using PCA, the feature importance can be seen in Figure \ref{fig:ShapleyWaterfallFirstTry}. The waterfall graph in Figure \ref{fig:ShapleyWaterfallFirstTry} shows that the Richardson number for higher levels is most important.

\begin{figure}[h]
    \caption{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
    \label{fig:ShapleyWaterfallFirstTry}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
\end{figure}

>>>>

\begin{table}[h]
    \centering
    \caption[Hyperparamter search with best performing combination.]{Hyperparamter search with best performing combination shown. Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on what is working and as such is neither exhaustive nor completely random. This means that an upper limit will not be set on the number of combinations to try like with randomsearch.}
    \label{table:gridSearchHyperparamters}
    \begin{tabular}{c|c|c}
        Parameter & Range of values & Selected\\\hline
        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\\hline
        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\\hline
        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\\hline
        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{30}\\\hline
        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\\hline
        Activation & ReLU, ELu, Softmax & \textbf{ReLU}\\\hline
    \end{tabular}
\end{table}

\section{Feature selection and importance}
Using the three datasources, a model is trained. The data from IMO only represents the ground truth and is not used as part of the training data. The gust ($f_g$) and wind speed ($f$) are used to calculate the target gust factor. All weather parameters used in predictions come from the CARRA reanalysis data, with variables wind speed, wind direction, temperature, pressure. All of these variables are queried in three height levels (15, 250, 500 meters above ground) at the location of a given weather station. The second data source for the training data, is the elevation data. Using the GeoTIFF file that IMO provided, a sector upwind were looked at.

\begin{figure}[h]
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
    \caption[Feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in table \ref{table:gridSearchHyperparamters} and data as described in table \ref{table:trainDataExample}.}
    \label{fig:ShapleyWaterfallFirstTry}
\end{figure}

### 2024-08-27 Chapters/Chapter5.tex
Updated after new CARRA data sourced and run

% Chapter Template

\chapter{Results} % Main chapter title

\label{Chapter5} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

%----------------------------------------------------------------------------------------
%	SECTION 1
%----------------------------------------------------------------------------------------

\section{Results}
A baseline model was constructed. This model looked at the gust factor for some training data and took the average of this and predicted this average everytime. The baseline model gave an average error of around \baselineerror\%. This sets a goal. A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline. Using the previously described neural network architecture a mean absolute percentage error of \modelerror\% was achived. This is some improvement upon the baseline error, with a decrease in error of around \errorImprovement\%.

\section{Discussion}
Looking only at the error, the performance of the model can be gauged. However it does not help us explain why the model made these predictions. Here feature interpretation can be of great help. Using Shapley values and the python SHAP package, the contribution of each attribute can be given a value, as previously mentioned. Figure \ref{fig:SHAP_final_model} shows the average contribution of each feature.

>>>>

\chapter{Results}
\label{Chapter5}
\section{Results}
A baseline model was constructed. This model looked at the gust factor for some training data and took the average of this and predicted this average everytime. Several baseline guesses were created based on a lower limit assigned to the average wind speed limit (AWSL). As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}. This means looking at a subset of the data where the AWSL is higher, a better result can be expected. The results show this. Several different AWSL were used for the baseline model, as can be seen in table \ref{table:results}. This sets a goal. A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline. Using the previously described neural network architecture setups for each AWSL, with and without landscape elevation information, MAPE was determined. The results can be seen in Table \ref{table:results}.

\begin{table}[h]
    \caption[Model results for different AWSL]{MAPE for each average wind speed limit with and without landscape elevation in a 30°sector around the point of interest into the direction of the reanalysis wind. The percentage error is higher for lower wind speeds and thus observing the error for different lower bound of wind speed will produce different results. How to determine this lower bound can done by measurements (f) or reanalysis wind speed at 15 m/s ($ws_{15}$). The influence of adding elevation data seems to reduce the error.}
    \label{table:results}
    \centering
    \begin{tabular}{c|cccccc}
        \textbf{AWSL} & \textbf{MAPE} & & & & &\\ 
        m/s & Baseline & &  Without elevation & & With elevation & \\
            & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\\hline
        0 & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
        5 & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
        7.5 & 25.3\% & 12.6\% & 14.5\% & 8.6\% & 12.5\% & 8.6\%\\
        10 & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
        12.5 & 23.3\% & 10.1\% & 13.1\% & 7.1\% & 11.0\% & 6.9\%\\
        15 & 23.2\% & 9.3\% & 13.0 & 7.0\% & 10.6\% & 6.4\%\\
        17.5 & 23.7\% & 8.7\% & 12.8\% & 6.4\% & 10.53\% & 6.1\%\\
        20 & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
        22.5 & 26.1\% & 7.8\% & 14.5\% & 6.1\% & 12.6\% & 5.7\%\\
        25 & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
    \end{tabular}
\end{table}

This is some improvement upon the baseline error, with a decrease in error of around \errorImprovement\%. The power generated by wind mill increases with wind speed cubed\cite{wind_power}. The highest wind gusts in Iceland are around 70 m/s. Knowing the gust factor to 2 percentage points better than before can allow for the anticipated power generated by wind mills to increase significantly.

\section{Discussion}
The error improvement is modest in absolute terms but represents 1.8\% over the baseline.

Training two identical models, one with the transformed reanalysis data and the other with original reanalysis data, results in two near identical errors, as expected. As no new information was input, the data was only manipulated, the model evaluates to the same error. The model that used the transformed reanalysis data was slightly quicker to converge.

Looking only at the error, the performance of the model can be gauged. However it does not help us explain why the model made these predictions. Here feature interpretation can be of great help. Using Shapley values and the python SHAP package, the contribution of each attribute can be given a value, as previously mentioned. Figure \ref{fig:SHAP_final_model} shows the average contribution of each feature.
